The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-07-05 paper/paper.Rnw
Fixed table and graphs. Everything ready.

\title{First paper - Draft}
\author{Jorge Cimentada}

@@

<<table_escs, include = TRUE>>=
summary_table <-
  left_join(summary_data, sample_tables_topbottom) %>%
  filter(name %in% c("2000", "2015")) %>%
  arrange(country, name)

table_coming <-
  summary_table %>%
  select(-perc) %>%
  gather(cats, vals, -(name:escs_dummy)) %>%
  unite(all_vals, escs_dummy, cats, sep = "_") %>%
  spread(all_vals, vals) %>%
  map_at(c(3, 4, 6, 7), ~ {
    .x[is.na(.x)] <- 0
    paste0(round(.x, 0), "%")
  }) %>%
  map_if(is_double, as.character) %>%
  data.frame() %>%
  xtable::xtable()

addtorow <- list()
addtorow$pos <- list(0)
addtorow$command <- paste("\\hline \\
                          & & & Low SES & & & High SES \\\\
                          \\cmidrule(l){3-5}
                          \\cmidrule(l){6-8}",
                          paste0(
                            c("Year",
                              "Countries",
                              "\\% Low Edu",
                              "\\% High Edu",
                              "Sample size",
                              "\\% Low Edu",
                              "\\% High Edu",
                              "Sample size"),
                            collapse = " & "),
                          "\\\\ \\hline")

@@

@@

us_math_2000_upper <-
  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
  .[["upper"]]
@
We find that the initial gap for the U.S in 2000 is \Sexpr{us_math_2000} varying between \Sexpr{us_math_2000_lower} and \Sexpr{us_math_2000_upper}. This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.05, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@
\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
\end{center}
\end{figure}

@@

<<graphing_achievement_disparity, cache = T>>=

@@

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between \Sexpr{max(correlation)} and \Sexpr{min(correlation)}. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.

>>>>

\title{The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}

@@

% <<table_escs, include = TRUE>>=
% summary_table <-
%   left_join(summary_data, sample_tables_topbottom) %>%
%   filter(name %in% c("2000", "2015")) %>%
%   arrange(country, name)
% 
% table_coming <-
%   summary_table %>%
%   select(-perc) %>%
%   gather(cats, vals, -(name:escs_dummy)) %>%
%   unite(all_vals, escs_dummy, cats, sep = "_") %>%
%   spread(all_vals, vals) %>%
%   map_at(c(3, 4, 6, 7), ~ {
%     .x[is.na(.x)] <- 0
%     paste0(round(.x, 0), "%")
%   }) %>%
%   map_if(is_double, as.character) %>%
%   data.frame() %>%
%   xtable::xtable()
% 
% addtorow <- list()
% addtorow$pos <- list(0)
% addtorow$command <- paste("\\hline \\
%                           & & & Low SES & & & High SES \\\\
%                           \\cmidrule(l){3-5}
%                           \\cmidrule(l){6-8}",
%                           paste0(
%                             c("Year",
%                               "Countries",
%                               "\\% Low Edu",
%                               "\\% High Edu",
%                               "Sample size",
%                               "\\% Low Edu",
%                               "\\% High Edu",
%                               "Sample size"),
%                             collapse = " & "),
%                           "\\\\ \\hline")
% 
% print(table_coming,
%       scalebox = '0.75',
%       floating = FALSE,
%       add.to.row = addtorow,
%       include.colnames = FALSE,
%       include.rownames = FALSE,
%       hline.after = nrow(table_coming),
%       caption.placement = "top",
%       caption = "SES sample size and ISCED composition")
% @

<<table_escs, include = TRUE>>=

avg_performance <-
  list(results_math[[1]], results_math[[6]]) %>%
  set_names(c(2000, 2015)) %>%
  enframe() %>%
  unnest()

summary_table <-
  sample_tables_topbottom %>%
  mutate(escs_dummy = as.character(escs_dummy)) %>%
  filter(name %in% c("2000", "2015")) %>%
  left_join(avg_performance) %>%
  arrange(country, name)

table_coming <-
  summary_table %>%
  select(-perc) %>%
  gather(cats, vals, -(name:escs_dummy)) %>%
  unite(all_vals, escs_dummy, cats, sep = "_") %>%
  spread(all_vals, vals) %>%
  select(name, country, `0_n`, `0_Mean`, `0_s.e.`, `1_n`, `1_Mean`, `1_s.e.`) %>%
  mutate(`0_n` = as.character(`0_n`),
         `1_n` = as.character(`1_n`)) %>%
  xtable::xtable()

addtorow <- list()
addtorow$pos <- list(0)
addtorow$command <- paste("\\hline \\
                          & & & Low SES & & & High SES \\\\
                          \\cmidrule(l){3-5}
                          \\cmidrule(l){6-8}",
                          paste0(
                            c("Year",
                              "Countries",
                              "N",
                              "Avg score",
                              "S.E",
                              "N",
                              "Avg score",
                              "S.E"),
                            collapse = " & "),
                          "\\\\ \\hline")

@@

The first table shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point. One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size. The table suggests that we have a reasonable number of respondents to actually estimate gaps accurately. Moreover, we can see that in all instances the bottom SES group has a lower score than all top SES groups.

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting their very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

@@

us_math_2000_upper <-
  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
  .[["upper"]]
@
We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}. This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

@@

full_data %>%
  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
  geom_rect(data = rect_data, aes(xmin = xst,
                                  xmax = xen,
                                  ymin = yst,
                                  ymax = yen),
            fill = rect_data$colour,
            alpha = 0.2,
            inherit.aes = FALSE) +
  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
            linetype = "longdash") +
  geom_point(alpha = 0.2) +
  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
                  aes(label = country), box.padding = unit(2.7, "lines")) +
  geom_vline(xintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  xlim(lims$xlim) +
  ylim(lims$ylim) +
  coord_cartesian(expand = FALSE) +
  annotate(geom = "text", x = 0.15, y = -0.2,
           label = "Low SES is catching up \n faster than High SES",
           fontface = 2, size = 3) +
  annotate(geom = "text", x = -0.075, y = 0.20,
           label = "High SES is increasing  \n faster than Low SES is catching up",
           fontface = 2, size = 3) +
  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
  theme_minimal()
@
\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
\end{center}
\end{figure}

@@

\subsection{Efficiency-equality tradeoff}
<<graphing_achievement_disparity, cache = T>>=

@@

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between \Sexpr{round(max(correlation), 2)} and \Sexpr{round(min(correlation), 2)}. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.
